---
title: "Erasmus+ Country-Level Mobility Analysis"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(sf)
library(scales)
library(readr)
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)

# Base directories 
proj_dir <- "your-path"
data_dir <- proj_dir
out_dir  <- file.path(proj_dir, "output")

```

## 1. Load raw Erasmus data

This script constructs a *country‑level* corridor dataset and runs all
models/aggregates used in the paper (H1 and H2). 

```{r load-data}
# Raw mobility data (one row per student)
erasmus_raw <- read_rds(file.path(data_dir, "erasmus_combined.rds"))

glimpse(head(erasmus_raw))
```

## 2. ISO codes and European analysis universe

We restrict to corridors where both sending and receiving countries are in the
European Erasmus+ programme/partner set and exclude within‑country moves.

```{r iso-europe-filter}
erasmus_raw <- erasmus_raw %>%
  rename(
    sending_city      = `Sending City`,
    receiving_city    = `Receiving City`,
    sending_country   = `Sending Country`,
    receiving_country = `Receiving Country`
  ) %>%
  mutate(
    send_code = str_extract(sending_country,   "^[A-Z]{2}"),
    recv_code = str_extract(receiving_country, "^[A-Z]{2}")
  )

# European analysis universe
europe_codes <- c(
  "AL","AT","BE","BG","CH","CY","CZ","DE","DK","EE","ES","FI",
  "FR","GR","HR","HU","IE","IS","IT","LI","LT","LU","LV","MT",
  "NL","NO","PL","PT","RO","SE","SI","SK","TR","UK","ME","MK",
  "RS","UA","MD","XK"
)

erasmus_raw <- erasmus_raw %>%
  filter(
    send_code %in% europe_codes,
    recv_code %in% europe_codes,
    send_code != recv_code
  )

```

## 2b. Population data (Eurostat-style, by country-year)

We read population and reshape to a country–year panel.

```{r population-data}
## 2b. Population data (Eurostat-style, by country-year)

pop_raw <- read_tsv(file.path(data_dir, "pop_count_.tsv"))

# First column: flags + country code, e.g. "A,NR,TOTAL,T,AT"
names(pop_raw)[1] <- "country_raw"

pop_long <- pop_raw %>%
  # Keep total population series
  filter(str_detect(country_raw, "A,NR,TOTAL,T")) %>%
  mutate(country = str_remove(country_raw, "A,NR,TOTAL,T,")) %>%
  select(country, `2014`:`2022`) %>%
  pivot_longer(
    cols      = `2014`:`2022`,
    names_to  = "year",
    values_to = "population"
  ) %>%
  mutate(
    year = as.numeric(year),
    # Strip flags and non-numeric chars, then convert
    population = gsub("[^0-9.,]", "", population),
    population = as.numeric(population)
  )

# Restrict to Erasmus years actually in the data
years_erasmus <- sort(unique(erasmus_raw$year))
pop_long <- pop_long %>%
  filter(year %in% years_erasmus)

# Restrict to our European universe
pop_long <- pop_long %>%
  filter(country %in% europe_codes)

cat("Population panel summary:\n")
summary(pop_long$population)
```

## 3. Country–country–year flows

```{r flows-country}
# Directed country–country–year flows
flows_country_raw <- erasmus_raw %>%
  count(send_code, recv_code, year, name = "n_students")

cat(sprintf("Country–country–year rows: %d\n", nrow(flows_country_raw)))
summary(flows_country_raw$n_students)
```

## 4. Country centroids, distances, and population attachment

We attach WGS84 centroids from Natural Earth, compute centroid‑to‑centroid
distances in km, and join sending/receiving population.

```{r centroids-distance}
# World countries
world_sf <- ne_countries(scale = "medium", returnclass = "sf")

# Harmonise problematic codes (e.g. GB → UK, -99 → iso_a2_eh)
world_sf <- world_sf %>%
  mutate(
    iso_a2 = if_else(iso_a2 == "-99", iso_a2_eh, iso_a2),
    iso_a2 = if_else(iso_a2 == "GB",  "UK", iso_a2)
  )

# Centroid coordinates
world_centroids <- world_sf %>%
  st_centroid() %>%
  mutate(
    lon = st_coordinates(.)[, 1],
    lat = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry() %>%
  select(iso_a2, lon, lat)

# Attach centroids to both ends of each corridor
flows_country_dist <- flows_country_raw %>%
  left_join(world_centroids, by = c("send_code" = "iso_a2")) %>%
  rename(lon_from = lon, lat_from = lat) %>%
  left_join(world_centroids, by = c("recv_code" = "iso_a2")) %>%
  rename(lon_to = lon, lat_to = lat) %>%
  filter(!is.na(lon_from), !is.na(lon_to))

cat(sprintf("Rows with both centroids: %d\n", nrow(flows_country_dist)))

# Great-circle distance (km)
sf_from <- st_as_sf(flows_country_dist,
                    coords = c("lon_from", "lat_from"), crs = 4326)
sf_to   <- st_as_sf(flows_country_dist,
                    coords = c("lon_to", "lat_to"), crs = 4326)

flows_country_dist$distance_km <- as.numeric(
  st_distance(sf_from, sf_to, by_element = TRUE)
) / 1000

summary(flows_country_dist$distance_km)
```

```{r attach-population}
# Attach population for sending and receiving countries (by ISO2 and year)
flows_country_dist <- flows_country_dist %>%
  left_join(pop_long, by = c("send_code" = "country", "year")) %>%
  rename(pop_sending = population) %>%
  left_join(pop_long, by = c("recv_code" = "country", "year")) %>%
  rename(pop_receiving = population)

cat("Population (sending) summary:\n")
summary(flows_country_dist$pop_sending)
cat("Population (receiving) summary:\n")
summary(flows_country_dist$pop_receiving)
```

## 5. Build corridor dataset and COVID periods

This is the main dataset for gravity models and COVID analysis.

```{r corridors-build}
corridors <- flows_country_dist %>%
  filter(
    !is.na(distance_km),
    !is.na(pop_sending), !is.na(pop_receiving),
    pop_sending > 0, pop_receiving > 0
  ) %>%
  mutate(
    log_n_students    = log(n_students),
    log_distance      = log(distance_km),
    log_pop_sending   = log(pop_sending),
    log_pop_receiving = log(pop_receiving),
    period = case_when(
      year %in% 2017:2019 ~ "pre_covid",
      year %in% 2020:2021 ~ "covid",
      year >= 2022        ~ "post_covid",
      TRUE                ~ "early"
    )
  )

```

## 6. Gravity models for H1 (distance and population)

We estimate:

- M1a: log–log gravity model with distance only.  
- M1b: distance + sending/receiving population (H1b).  
- M1c: distance + population + year fixed effects.

```{r gravity-models}
# M1a: distance only
m1a <- lm(log_n_students ~ log_distance, data = corridors)

# M1b: distance + sending/receiving population
m1b <- lm(log_n_students ~ log_distance + log_pop_sending + log_pop_receiving,
          data = corridors)

# M1c: distance + population + year FE
m1c <- lm(log_n_students ~ log_distance + log_pop_sending + log_pop_receiving +
            factor(year),
          data = corridors)

model_list   <- list(m1a, m1b, m1c)
model_labels <- c("M1a: distance only",
                  "M1b: + population",
                  "M1c: + population + year FE")

for (i in seq_along(model_list)) {
  m  <- model_list[[i]]
  se <- sqrt(diag(vcov(m)))
  cat(sprintf(
    "\n%s\n  Distance coef: %.3f (SE = %.3f) | Adj.R² = %.3f\n",
    model_labels[i],
    coef(m)["log_distance"],
    se["log_distance"],
    summary(m)$adj.r.squared
  ))
}

summary(m1c)
```

## 7. Peripheral senders vs core (H1c)

We test whether selected geographically peripheral senders tend to have longer
corridor distances.

```{r periphery-model}
# Theory-driven periphery definition
periphery <- c("IS","IE","CY","PT","FI","NO","SE","MT")

corridors <- corridors %>%
  mutate(peripheral_sender = send_code %in% periphery)

m_periph <- lm(distance_km ~ peripheral_sender + factor(year), data = corridors)

summary(m_periph)
```

## 8. COVID disruption (H2a–H2c)

### 8.1 Total flows by period (H2a)

```{r total-by-period}
total_by_period <- corridors %>%
  group_by(period) %>%
  summarise(
    total_flows = sum(n_students),
    mean_dist   = mean(distance_km),
    .groups     = "drop"
  )

total_by_period

# Simple log-linear comparison of periods (COVID as reference)
m_period <- lm(log(total_flows) ~ period, data = total_by_period)
summary(m_period)
```

### 8.2 Distance bands by period (H2b)

```{r distance-bands}
corridors_distcat <- corridors %>%
  mutate(
    dist_cat = case_when(
      distance_km <= 1000 ~ "short",
      distance_km <= 2500 ~ "medium",
      TRUE                ~ "long"
    )
  )

flows_by_period_dist <- corridors_distcat %>%
  group_by(period, dist_cat) %>%
  summarise(total = sum(n_students), .groups = "drop") %>%
  group_by(period) %>%
  mutate(share = total / sum(total)) %>%
  ungroup()

flows_by_period_dist
```

```{r plot-distance-bands}
p_bands <- ggplot(flows_by_period_dist,
                  aes(x = period, y = share, fill = dist_cat)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_viridis_d(name = "Distance band") +
  theme_minimal() +
  labs(
    title = "Share of flows by distance band and period",
    x     = NULL,
    y     = "Share of flows"
  )

p_bands

ggsave(file.path(out_dir, "erasmus_flows_distance_bands_period.png"),
       plot = p_bands, dpi = 300, width = 8, height = 5)
```

### 8.3 Top‑10 corridor concentration by period (H2c)

```{r top10-share}
top10_share <- corridors %>%
  group_by(period, send_code, recv_code) %>%
  summarise(n = sum(n_students), .groups = "drop") %>%
  group_by(period) %>%
  mutate(rank = rank(-n, ties.method = "first")) %>%
  summarise(
    total_all   = sum(n),
    top10_total = sum(n[rank <= 10]),
    top10_share = top10_total / total_all,
    .groups     = "drop"
  )

top10_share
```

```{r plot-top10-share}
p_top10 <- ggplot(top10_share, aes(x = period, y = top10_share)) +
  geom_line(group = 1, colour = "#d7191c") +
  geom_point(size = 3, colour = "#d7191c") +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  theme_minimal() +
  labs(
    title = "Top‑10 corridor share of total flows by period",
    x     = NULL,
    y     = "Share of total flows"
  )

p_top10

ggsave(file.path(out_dir, "erasmus_top10_share_period.png"),
       plot = p_top10, dpi = 300, width = 8, height = 5)
```

