---
title: "Erasmus+ Participation vs GDP - Basic Country Statistics"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(max.print = 75)

library(tidyr)
library(scales)
library(dplyr)
library(tidyverse)
library(readr)
library(lme4)
library(ggplot2)
library(stringr)
library(MuMIn)
library(sf)
library(tmap)
library(spdep)
library(mapview)
library(purrr)
library(cartogram)
library(spatialreg)

# Base directories (align with other scripts)
proj_dir <- "your-path"
data_dir <- proj_dir
out_dir  <- file.path(proj_dir, "output")

```

## 1. Erasmus+ data (country–year participation)

```{r erasmus-data}
data <- read_rds(file.path(data_dir, "erasmus_combined.rds"))

# Drop first column if it is an index
data <- data[, -1]

data <- data %>%
  rename(
    sending_city     = `Sending City`,
    receiving_city   = `Receiving City`,
    sending_country  = `Sending Country`,
    receiving_country = `Receiving Country`
  ) %>%
  mutate(
    send_code = str_extract(sending_country,  "^[A-Z]{2}"),
    recv_code = str_extract(receiving_country, "^[A-Z]{2}")
  )

# European analysis universe 
europe_codes <- c(
  "AL","AT","BE","BG","CH","CY","CZ","DE","DK","EE","ES","FI",
  "FR","GR","HR","HU","IE","IS","IT","LI","LT","LU","LV","MT",
  "NL","NO","PL","PT","RO","SE","SI","SK","TR","UK","ME","MK",
  "RS","UA","MD","XK"
)

data <- data %>%
  filter(
    send_code %in% europe_codes,
    recv_code %in% europe_codes,
    send_code != recv_code
  )

cat(sprintf("Rows after European filter: %d\n", nrow(data)))
```

## 2. GDP and population data 

We load GDP and population files and build a country–year panel.

```{r gdp-data}
gdp_data <- readRDS(file.path(data_dir, "gdp_data.rds"))
gdp_data <- gdp_data %>%
  rename(
    country   = geo,
    year      = TIME_PERIOD,
    gdp_value = OBS_VALUE
  )
```

```{r pop-data}
pop_data <- read_tsv(file.path(data_dir, "pop_count_.tsv"))

# First column contains flags + country code, e.g. "A,NR,TOTAL,T,AT"
names(pop_data)[1] <- c('country')

pop_data <- pop_data %>%
  filter(str_detect(country,"A,NR,TOTAL,T"))
pop_data$country <- str_remove(pop_data$country,"A,NR,TOTAL,T,")

pop_data <- pop_data[, c(1, 56:64)]
```

```{r pop-long}
pop_data_long <- pop_data %>%
  pivot_longer(
    cols      = `2014`:`2022`,
    names_to  = "year",
    values_to = "population"
  ) %>%
  mutate(year = as.numeric(year))
```

```{r merge-pop-gdp}
add_df <- pop_data_long %>%
  inner_join(gdp_data, by = c("country", "year"))

# Check coverage
print(sort(unique(add_df$year)))
```

## 3. Restrict to Erasmus countries and clean numeric fields

```{r filter-to-erasmus}
# Countries actually appearing in Erasmus flows (send or receive)
erasmus_countries <- data %>%
  pivot_longer(cols = c(send_code, recv_code),
               names_to = "type",
               values_to = "country") %>%
  distinct(country) %>%
  pull(country)

add_df <- add_df %>%
  filter(country %in% erasmus_countries)

# Keep only digits, comma, dot; then convert to numeric
add_df$gdp_value  <- gsub("[^0-9.,]", "", add_df$gdp_value)
add_df$population <- gsub("[^0-9.,]", "", add_df$population)

add_df$gdp_value  <- as.numeric(add_df$gdp_value)
add_df$population <- as.numeric(add_df$population)

summary(add_df$gdp_value)
summary(add_df$population)
```

```{r manual-fix}
# Manual correction for two missing population values 
# These numbers correspond to specific (country, year) entries inspected beforehand.
add_df$population[340:341] <- c(1793000, 1745000)

# Simple GDP-per-capita proxy (GDP / population * 100 for nicer magnitudes)
add_df$norm_gdp_val <- round((add_df$gdp_value / add_df$population) * 100, 2)

summary(add_df$norm_gdp_val)
```

## 4. Build country–year participation table

We aggregate Erasmus sends by country and year and attach population and GDP.

```{r df-country}
# Rename for join
names(add_df) <- c("send_code", "year", "population", "gdp_value", "norm_gdp_val")

# Erasmus participants sent per country-year
df_country <- data %>%
  count(send_code, year, name = "n_participants")

# Unique population/GDP per (send_code, year)
df_pop <- add_df %>%
  distinct(send_code, year, population, gdp_value, norm_gdp_val)

df_country <- df_country %>%
  inner_join(df_pop, by = c("send_code", "year"))

# For convenience
df_country <- df_country %>%
  rename(CNTR_ID = send_code)

glimpse(df_country)
```

```{r participation-and-gdp}
# Participants per population
df_country$participation_rate <- ((df_country$n_participants /df_country$population)*100)
```

```{r}
# Add geospatial data
polygon_data <- st_read('country_borders.gpkg')
df_country$CNTR_ID <- str_split_fixed(df_country$CNTR_ID, " ", 2)[,1]

geo_data <- polygon_data %>% 
  inner_join(df_country, by=c("CNTR_ID"))
```

## 5. Plot the raw participation rates

```{r}
# plotting the raw participation rate for EU countries between the years 2019-2022
geo_data[geo_data$year %in% c("2019","2020","2021","2022"), ] %>%
  tm_shape() +
  tm_facets(by = "year") +
  tm_polygons(
    fill = "participation_rate",
    palette = c("#F67451", "#E6D460", "#00E078", "#73E7D8"),
    title = "Participation frequency"
  ) +
  tm_layout(
    legend.outside = TRUE,
    legend.outside.position = "bottom"
  ) +
  tm_tiles()

# Plotting proportional participation map
df_carto_all <- geo_data %>%
  split(.$year) %>%
  map(~ cartogram_cont(.x, weight = "participation_rate", itermax = 5)) %>%
  bind_rows()

ggplot(df_carto_all) +
  geom_sf(aes(fill = participation_rate)) +
  facet_wrap(~ year) +
  scale_fill_gradientn(
    colours = c("#064789", "#8CD790", "#FCFDBF", "#FE9F6D", "#DE4968"),
    values = scales::rescale(c(0, 0.25, 0.5, 0.75, 1))
  ) +
  labs(title="Proportional share of participation in Erasmus+", fill="share of participants")+
  theme_minimal()
```

## 7. Analyse the autocorelation

```{r}
# Calculate the queen contiguity neighbourhoods
nb <- poly2nb(geo_data, queen = TRUE)
country_nb <- st_coordinates(st_centroid(geo_data$geom))

lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Moran test for GDP and participation rate with Monte-Carlo simulation
moran.mc(geo_data$participation_rate, 
           lw,
           nsim = 999,
           zero.policy = TRUE)
moran.mc(geo_data$norm_gdp_val, 
           lw,
           nsim = 999,
           zero.policy = TRUE)
```

```{r}
# Calculte univ LISA clusters for participation rate and the bv LISA clusters for participation and GDP
lisa_participation <- localmoran(geo_data$participation_rate,lw,
            na.action=na.fail,
        conditional=TRUE, alternative = "two.sided", mlvar=TRUE,
        spChk=NULL, adjust.x=FALSE, zero.policy=attr(lw, "zero.policy"))

bv_lisa_participation_GDP <- localmoran_bv(geo_data$participation_rate, geo_data$norm_gdp_val,
           lw,nsim = 499, scale = TRUE, alternative="two.sided",
 iseed=1L, no_repeat_in_row=FALSE, zero.policy=attr(lw, "zero.policy"))
```

## 8. Plot LISA clusters

```{r}
# Plotting univ LISA clusters for participation rate and the bv LISA clusters for participation and GDP
geo_data$lisa_uni_cluster <- attr(lisa_participation, "quadr")[,'pysal']
geo_data$lisa_bv_cluster <- attr(bv_lisa_participation_GDP, "quadr")[,'pysal']

# univ LISA
tm_shape(geo_data)+
  tm_polygons("lisa_uni_cluster",
              palette = c("#064789", "#8CD790", "#FE9F6D", "#DE4968"),
              title = "LISA Clusters") +
  tm_layout(frame = FALSE)
# bv LISA
tm_shape(geo_data)+
  tm_polygons("lisa_bv_cluster",
              palette = c("#064789", "#8CD790", "#FE9F6D", "#DE4968"),
              title = "LISA Clusters") +
  tm_layout(frame = FALSE)
```

## 9. Build spatial models 

```{r}
# apply the SLX model to the data
GDP_participation_model01 <- lmSLX(participation_rate ~ norm_gdp_val + year, data = geo_data, listw = lw)
summary(GDP_participation_model01)
```

```{r}
# apply the OLS and the SEM model to the data
GDP_participation_model02 <- errorsarlm(participation_rate ~ norm_gdp_val + year, data = geo_data, listw = lw)
summary(GDP_participation_model02)
```

